\textbf{Political Instability and Financial Markets}

### 2017-07-09 doc/paper.tex
Add empty data directory

\singlespacing
\begin{abstract}
We examine the relationship between political instability and the daily returns of national stock indices. Financial volatility increases dramatically following (and just before) ``irregular'' regime changes caused by coup d'état, assassinations, and resignations. Some of the pre-resignation volatility occurs because of public protests such as riots and popular demonstrations that helped spawn the resignation in the first place. However, these irregular regime changes have disparate effects on the direction of stock returns. Abnormal returns following resignations are large and positive (6\%) while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events such as the temporary ousting of Hugo Chavez during the 2002 Venezuelan coup d'état attempt result in positive abnormal returns of 20\% or more.
\end{abstract}
\doublespacing

@@

My main finding is that financial volatility increases dramatically following (and just before) coup d'état, assassinations, and resignations. However, these irregular regime changes have disparate effects on the direction of stock returns.  Abnormal returns following resignations are large and positive (6\%) while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events such as the temporary ousting of Hugo Chavez during the 2002 Venezuelan coup d'état attempt result in positive abnormal returns of 20\% or more. We also present evidence showing that volatility increases during public protests such as riots and popular demonstrations that cause leaders to resign. These findings provide important insights into the expected impact of episodes of political instability on economic performance.

@@

The fourth looks at how the variance of stock market returns reacts to political events. \citet{leblang2005government} show that right-wing governments increase the mean and volatility of stock prices in the United States and Britain. \citet{jensen2005market} find that a rise in the popularity of a Brazilian presidential candidate with uncertain future policies, Luiz Inácio Lula da Silva, increased the volatility (but not the mean return) of the Brazilian stock market.

\section{Data}
Political data are primarily drawn from the Center for Systemic Peace's (CSP) Polity IV Coup d'état dataset and Coup d'état Events handbook. The Coup d'état dataset incudes the date of 1) successful coups, 2) attempted coups, 3) plotted coups and 4) alleged coup plots. We focus on successful coups because it is difficult to classify failed coups. \citet[p. 617]{needler1966political} has even gone so far as to say that ``the categories of coups that were aborted, suppressed, or abandoned melt into each other and into a host of other non-coup phenomena so as to defy accounting.''

The Coup d'état Events handbook provides a list of 1) auto-coups, 2) the ouster of leadership by foreign forces, 3) the ouster of leadership by rebel forces, 4) assassinations of the executive and 5) resignations of the executive due to poor performance and/or loss of authority. Daily financial data is available for countries in categories 4 and 5, so We supplement the coups with assassinations and resignations to form a list of ``irregular'' regime changes. The resignations are those in which the ruling executive was coerced to resign due to poor performance, public discontent and popular demonstrations.

@@

Financial data are from the Global Financial Data database, which includes the longest available time series of stock prices. We collect data on national equity indices and two global equity indices, the S\&P/IFC Emerging Market Investable Composite and the Morgan Stanley Capital International (MSCI) World Index. The S\&P/IFC index includes securities from emerging markets while the MSCI index includes securities from developed markets only. We collect stock index data on every country in which there was a coup or coup attempt, an assassination or failed assassination, or a forced resignation.\footnote{The list of failed assassinations are from \citet{jones2009hit}. Coup attempts are those in category 2 in the CSP Coup d'état dataset.}The longest available time series for these stock indices are listed in \autoref{tab:stock list}.

@@

We define the event date as the first trading day in which the market could have reacted to news of the event. For example, during the October 12, 1999 coup d'état in Pakistan led by General Pervez Musharraf, the army announced that Prime Minister Nawaz Sharif had been dismissed after market hours at 10:15 pm. We code October 14th, the day in which the market re-opened, as the event day. When events occurred on weekends, We change the event date to the following Monday.

@@

We report abnormal returns separately for coups, assassinations and resignations because they may have distinct effects on stock returns. \autoref{tab:AR-coups} shows abnormal returns for national stock indices both preceding and following coup d'état.
Standard errors and p-values are calculated using asymptotic t-statistics as in \citet{mackinlay1997event}. \footnote{It is appropriate to use the standard normal distribution to calculate test statistics because the length of the estimation window is sufficiently long (250 trading days).}

@@

Although it is tempting to dismiss the 1976 Argentina coup as an outlier, there is reason to to believe that coups can in fact result in large positive abnormal returns. The 2002 failed coup d'état attempt against Hugo Chavez is a case in point. On the evening of April 11, 2002, coup plotters removed Chavez from office and later detained him. Pedro Carmona, a Venezuelan business leader, was named the transitional President of Venezuela. Two days later, on April 13, 2002, a popular uprising led to Chavez's reinstatement as president.

@@

\begin{figure}[!ht]
\includegraphics{../figs/venezuela_coup_attempt_2002.pdf}
\caption{Abnormal returns surrounding the 2002 Venezuelan coup d'état attempt}
\label{fig:AR-Ven}
\end{figure}

@@

The third concern, which is that the political events in this paper are not unexpected, is addressed in a number of ways. First, one can note that neither the 7-day or the 20-day mean CAR from tables \autoref{tab:AR-coups}, \autoref{tab:AR-ass} or \autoref{tab:AR-resignations} are statistically different than zero at even the 10\% level. Although some of the irregular regime changes have statistically significant pre-event CAR's, most of them do not, and the ones that do do not always move in the same direction. For example, there was a negative 11.8\% 7-day CAR preceding the resignation of President Estrada of the Philippines in January of 2001 and a positive 14.7\% 7-day CAR preceding the resignation of President Fernando de la Rúa of Argentina in December of 2001, even though both resignations were associated with large positive event day abnormal returns.

@@

An important question is whether other popular uprisings have had similar adverse economic consequences. To examine this, we explore all resignations that were driven by significant public protests.\footnote{The set of resignations includes all those listed in either the Coup d'état Events Handbook or the Archigos Version 2.9 data set with available financial data. In practice, this is the 2011 Egyptian Revolution and the list or resignations in \autoref{tab:AR-resignations}.} Public protests include popular demonstrations, riots, non-violent civil resistance and other forms of public discontent. These events are listed in \autoref{tab:protest-list} in the appendix.

The start and end dates in \autoref{tab:protest-list} are the dates that protests began and leader's resigned respectively. Resignations caused by popular uprisings were identified by examining the descriptions in the Coup d'état Events Handbook and Archigos Version 2.9. Additional Lexis Nexis searches were used to verify these descriptions. Start dates were determined through similar means.

>>>>

\singlespacing
\begin{abstract}
We examine the relationship between political instability and the daily returns of national stock indices. Financial volatility increases dramatically following (and just before) ``irregular'' regime changes caused by coup d'?tat, assassinations, and resignations. Some of the pre-resignation volatility occurs because of public protests such as riots and popular demonstrations that helped spawn the resignation in the first place. However, these irregular regime changes have disparate effects on the direction of stock returns. Abnormal returns following resignations are large and positive (6\%) while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events such as the temporary ousting of Hugo Chavez during the 2002 Venezuelan coup d'?tat attempt result in positive abnormal returns of 20\% or more.
\end{abstract}
\doublespacing

@@

My main finding is that financial volatility increases dramatically following (and just before) coup d'?tat, assassinations, and resignations. However, these irregular regime changes have disparate effects on the direction of stock returns.  Abnormal returns following resignations are large and positive (6\%) while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events such as the temporary ousting of Hugo Chavez during the 2002 Venezuelan coup d'?tat attempt result in positive abnormal returns of 20\% or more. We also present evidence showing that volatility increases during public protests such as riots and popular demonstrations that cause leaders to resign. These findings provide important insights into the expected impact of episodes of political instability on economic performance.

@@

The fourth looks at how the variance of stock market returns reacts to political events. \citet{leblang2005government} show that right-wing governments increase the mean and volatility of stock prices in the United States and Britain. \citet{jensen2005market} find that a rise in the popularity of a Brazilian presidential candidate with uncertain future policies, Luiz In?cio Lula da Silva, increased the volatility (but not the mean return) of the Brazilian stock market.

\section{Data}
Political data are primarily drawn from the Center for Systemic Peace's (CSP) Polity IV Coup d'?tat dataset and Coup d'?tat Events handbook. The Coup d'?tat dataset incudes the date of 1) successful coups, 2) attempted coups, 3) plotted coups and 4) alleged coup plots. We focus on successful coups because it is difficult to classify failed coups. \citet[p. 617]{needler1966political} has even gone so far as to say that ``the categories of coups that were aborted, suppressed, or abandoned melt into each other and into a host of other non-coup phenomena so as to defy accounting.''

The Coup d'?tat Events handbook provides a list of 1) auto-coups, 2) the ouster of leadership by foreign forces, 3) the ouster of leadership by rebel forces, 4) assassinations of the executive and 5) resignations of the executive due to poor performance and/or loss of authority. Daily financial data is available for countries in categories 4 and 5, so We supplement the coups with assassinations and resignations to form a list of ``irregular'' regime changes. The resignations are those in which the ruling executive was coerced to resign due to poor performance, public discontent and popular demonstrations.

@@

Financial data are from the Global Financial Data database, which includes the longest available time series of stock prices. We collect data on national equity indices and two global equity indices, the S\&P/IFC Emerging Market Investable Composite and the Morgan Stanley Capital International (MSCI) World Index. The S\&P/IFC index includes securities from emerging markets while the MSCI index includes securities from developed markets only. We collect stock index data on every country in which there was a coup or coup attempt, an assassination or failed assassination, or a forced resignation.\footnote{The list of failed assassinations are from \citet{jones2009hit}. Coup attempts are those in category 2 in the CSP Coup d'?tat dataset.}The longest available time series for these stock indices are listed in \autoref{tab:stock list}.

@@

We define the event date as the first trading day in which the market could have reacted to news of the event. For example, during the October 12, 1999 coup d'?tat in Pakistan led by General Pervez Musharraf, the army announced that Prime Minister Nawaz Sharif had been dismissed after market hours at 10:15 pm. We code October 14th, the day in which the market re-opened, as the event day. When events occurred on weekends, We change the event date to the following Monday.

@@

We report abnormal returns separately for coups, assassinations and resignations because they may have distinct effects on stock returns. \autoref{tab:AR-coups} shows abnormal returns for national stock indices both preceding and following coup d'?tat.
Standard errors and p-values are calculated using asymptotic t-statistics as in \citet{mackinlay1997event}. \footnote{It is appropriate to use the standard normal distribution to calculate test statistics because the length of the estimation window is sufficiently long (250 trading days).}

@@

Although it is tempting to dismiss the 1976 Argentina coup as an outlier, there is reason to to believe that coups can in fact result in large positive abnormal returns. The 2002 failed coup d'?tat attempt against Hugo Chavez is a case in point. On the evening of April 11, 2002, coup plotters removed Chavez from office and later detained him. Pedro Carmona, a Venezuelan business leader, was named the transitional President of Venezuela. Two days later, on April 13, 2002, a popular uprising led to Chavez's reinstatement as president.

@@

\begin{figure}[!ht]
\includegraphics{../figs/venezuela_coup_attempt_2002.pdf}
\caption{Abnormal returns surrounding the 2002 Venezuelan coup d'?tat attempt}
\label{fig:AR-Ven}
\end{figure}

@@

The third concern, which is that the political events in this paper are not unexpected, is addressed in a number of ways. First, one can note that neither the 7-day or the 20-day mean CAR from tables \autoref{tab:AR-coups}, \autoref{tab:AR-ass} or \autoref{tab:AR-resignations} are statistically different than zero at even the 10\% level. Although some of the irregular regime changes have statistically significant pre-event CAR's, most of them do not, and the ones that do do not always move in the same direction. For example, there was a negative 11.8\% 7-day CAR preceding the resignation of President Estrada of the Philippines in January of 2001 and a positive 14.7\% 7-day CAR preceding the resignation of President Fernando de la R?a of Argentina in December of 2001, even though both resignations were associated with large positive event day abnormal returns.

@@

An important question is whether other popular uprisings have had similar adverse economic consequences. To examine this, we explore all resignations that were driven by significant public protests.\footnote{The set of resignations includes all those listed in either the Coup d'?tat Events Handbook or the Archigos Version 2.9 data set with available financial data. In practice, this is the 2011 Egyptian Revolution and the list or resignations in \autoref{tab:AR-resignations}.} Public protests include popular demonstrations, riots, non-violent civil resistance and other forms of public discontent. These events are listed in \autoref{tab:protest-list} in the appendix.

The start and end dates in \autoref{tab:protest-list} are the dates that protests began and leader's resigned respectively. Resignations caused by popular uprisings were identified by examining the descriptions in the Coup d'?tat Events Handbook and Archigos Version 2.9. Additional Lexis Nexis searches were used to verify these descriptions. Start dates were determined through similar means.
